---
title: "Is there evidence of White Flight in Boston over the last 20 years?"
author: Rachel Kwon, Erin Mahoney, & Kevin Nazario
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    theme: yeti
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard) # for the dashboard
library(shiny)
library(magrittr)
library(stringr)
library(RColorBrewer)
library(viridis)

#map libraries - need these to load shapefile
library(leaflet)
library(htmltools)
library(tigris)
library(sf)
load('Saved.Rdata') # load data & plots from same directory as RMD file


# color choice variables into vector
color_choices <- c("Viridis"= "viridis", "Magma"="magma", "Plasma"="plasma", "Inferno"="inferno", "Spectral"= "Spectral", "Red-Yellow-Blue"= "RdYlBu", "Blues"= "Blues", "Random"= "Set1")

#set city variables into vector
cities <- c('Boston, MA'='MA', 'D.C.' = 'DC', 'Bay Area, CA'= 'CA', 'Seattle, WA' = 'WA', 'Portland, OR'= 'OR', 'Knoxville, TN'= 'TN', 'Detroit, MI' = 'MI') 

```

Maps{data-orientation=rows}
=====================================

Column {.sidebar} 
-----------------------------------------------------------------------

### Visualizing Change in White population over time

```{r}
#City Dropdown
selectInput("cities", "Choose a city:",
      choices = cities, 
      selected = cities[1])
#Year Dropdown
selectInput("year", "Choose a year:",
      c('1990', '2000', '2010'), 
      selected = '1990')
#Color Drop Down
selectInput("palette", "Choose a color palette:",
            choices = color_choices, selected = color_choices[1])

checkboxInput("reverse", label = "Reverse Colors?")
```

**Note: The map will update every time you change the city, year or color scheme. 

Hovering over the map will show the percentage of white people in the current tract as well as the number of white people (i.e. 35% (200) means the tract is 35% white with 200 white people according to the decennial census). 

Clicking on an area of the map will tell you the name/number of the given tract. 

Row
-----------------------------------------------------------------------

### 
```{r} 
#Render map
(map1<- renderLeaflet({                 
  
  #assigning appropriate dataset based on city and year input
  tmp <- get(paste('cendat_',input$cities,"_",input$year, sep = "")) 
  pct_white <- tmp$pct_white #assigning appropriate percent white value 
  
  #assinging appropriate number of white based on year (variable changes based on census year)
  if (input$year == 1990){
  num_white <- tmp$P0070001
} else if (input$year ==2000){
  num_white <- tmp$P003003
}else{
  num_white <-tmp$P003002 # 2010
}
 #set color pallate with range 0-100 for full spectrum 
   pal_all = colorNumeric(palette = input$palette, domain = c(0,100))
 if (input$reverse){
   pal_all = colorNumeric(palette = input$palette, domain = c(0,100), reverse = TRUE)
 }
  
  if(nrow(tmp) >= 1){
    tmp %>% st_transform(crs = "+init=epsg:4326")%>%
    leaflet(height = '100%') %>%  
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"), # click on tract for tract name/number 
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.5,
              color = ~ pal_all(pct_white),
              label = ~paste0(pct_white, "%, (", num_white, ")")) %>% # hover over tract for % and number
    addLegend("bottomright", 
            pal = pal_all, 
            values = ~ pct_white,
            title = "Census Tract Pct White",
            opacity = 1)
  } else leaflet()
}))

```

Comparing Maps
=====================================
Column {.sidebar}
-----------------------------------------------------------------------

### Comparing White population change over time

```{r}
selectInput("cities_compare", "Choose a city:",
      choices = cities, 
      selected = cities[1])

selectInput("year_left", "Left Map Year:",
      c('1990', '2000', '2010'), 
      selected = '1990')

selectInput("year_right", "Right Map Year:",
      c('1990', '2000', '2010'), 
      selected = '2000')


selectInput("palette2", "Choose a color palette:",
            choices = color_choices, selected = color_choices[1])

checkboxInput("reverse2", label = "Reverse Colors?")
```

Note: The map will update every time you change the city, year or color scheme. 

You can compare the same city across years by changing the Left and Right Map Year options

Changing the city or the color will update both maps to match. 

Hovering over the map will show the percentage of white people in the current tract as well as the number of white people (i.e. 35% (200) means the tract is 35% white with 200 white people according to the decennial census). 

Clicking on an area of the map will tell you the name/number of the given tract. 

Column
-------------------------------------

### 

```{r}
(map2<- renderLeaflet({                 
  
  tmp_left <- get(paste('cendat_',input$cities_compare,"_",input$year_left, sep = ""))
  pct_white_left <- tmp_left$pct_white   
  pal_all = colorNumeric(palette = input$palette2, domain = c(0,100))
  
  if (input$reverse2){
    pal_all = colorNumeric(palette = input$palette2, domain = c(0,100), reverse = TRUE)
 }
  
  if (input$year_left == 1990){
    num_white_left  <- tmp_left$P0070001
} else if (input$year_left ==2000){
    num_white_left <- tmp_left$P003003
} else{
    num_white_left <-tmp_left$P003002 # 2010 
}
  
  if(nrow(tmp_left) >= 1){
    tmp_left %>% st_transform(crs = "+init=epsg:4326")%>%
    leaflet() %>%  
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.5,
              color = ~ pal_all(pct_white_left),
              label = ~paste0(pct_white_left, "%, (", num_white_left, ")"))
  } else leaflet()
}))

```

Column
-------------------------------------
   
### 

```{r}

(map3<- renderLeaflet({                 

  tmp_right <- get(paste('cendat_',input$cities_compare,"_",input$year_right, sep = ""))
  pct_white_right <- tmp_right$pct_white   
  pal_all2 = colorNumeric(palette = input$palette2, domain = c(0,100))

  if (input$reverse2){
    pal_all2 = colorNumeric(palette = input$palette2, domain = c(0,100), reverse = TRUE)
  }
  
  if (input$year_right == 1990){
    num_white_right  <- tmp_right$P0070001
} else if (input$year_right ==2000){
    num_white_right <- tmp_right$P003003
} else{
    num_white_right <-tmp_right$P003002 # 2010 
}
 
  if(nrow(tmp_right) >= 1){
    tmp_right %>% st_transform(crs = "+init=epsg:4326")%>%
    leaflet() %>%  
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(NAME, "^([^,]*)"),
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.5,
              color = ~ pal_all2(pct_white_right),
              label = ~paste0(pct_white_right, "%, (", num_white_right, ")")) %>%
    addLegend("bottomright", 
            pal = pal_all2, 
            values = c(0,100),
            title = "Census Tract Pct White",
            opacity = 1)
  } else leaflet()
}))
```   

Permutation Test
=====================================
Column {.tabset .tabset-fade}
-------------------------------------
   
### Figure X
```{r}
plot1
```   
 
### Figure Y 
```{r}
plot2
```

### Figure Z 
```{r}
plot3
```

Column 
-------------------------------------
    
### Analysis
We decided to permute the years for the city of Boston to test the null hypothesis without making any assumptions about the distribution. The permutation test is important because it does not rely on assumptions such as normality by resampling the data many times in order to find a p-value for the permutation distribution. Thus, three permutation tests were performed for the following:

 1. Permuting 1990 & 2000
 2. Permuting 2000 & 2010
 3. Permuting 1990 & 2010

If the null hypothesis is true, there changing the sampling of the year should have **no effect** on the outcome, percentage of white people in the census tracts. 					

For the permutation test, we decided to create 10000 permutations of the assigned conditions. Then, for each of these events, we took the mean difference between the two years for that assignment. Next, we compared the distribution of mean differences we found through the permutation sampling to the observed mean difference. **Figure X** shows the distribution of mean differences from the permutation test and the vertical line is the observed mean difference in affect score between the years 1990 and 2000. **Figure Y** is the distribution of permuted mean differences between the years 2000 & 2010. **Figure Z** is the distribution of permuted mean differences between the years 1990 and 2010. For all three of the permutation tests, we can see that **we can successfully reject the null hypothesis** that there are no observed differences across the years for the percentage of white people in the city of Boston. However, we can see that over time, the observed mean difference is increasingly further away from the permuted distribution of mean differences, which can be seen in Figure Z. This confirms our hypothesis that over time, there have been changes in the percentage of white people in Boston. 
```{r}
```

The approximated **p-value** based on the permutation test was 

Project Narrative/Write-Up
=====================================
Column 
------------------------------------- 
### 

**The original question:**

We began this final project with an interest in population demographic data and creating visualizations, and settled on a goal: can we visualize racial population movement (a subset of which is often referred to as "white flight") over time, particularly in urban areas?

**Dataset**

To understand population demographics and movement over time, we considered using either the decennial census or the (annual) American Community Survey, and spent a considerable amount of time exploring the data available from each. We ultimately concluded that the ACS, which admittedly provides more granular data on a time scale, would likely not provide us any meaningful data about changes over time. ACS data is available by API for only a few years within the last decade, whereas the decennial census allows us to gather data spanning multiple decades. Furthermore, ACS data takes a population sample and gathers data about them to extrapolate to the full United States population, whereas the decennial census attempts to tabulate population numbers, a feature which we found more directly applicable to our goals.

**Timescale**

The decennial census is available by API for three rollouts of the census: 1990, 2000, and 2010, so we chose to use these three years of data. As a side note, we're guessing that the fact these are the only data available may be related to the fact that the 1990 census was the first one in which the entirety of the country was broken into census tracts; in previous census rollouts, census tracts covered only urban areas. 

**Geometry**

Although initially we began collecting and displaying data by zip code, with the rationale that zip codes are generally publicly known and easily understood, we ultimately chose to gather and display data by census tract. Tracts are drawn by the Census Bureau and meant to be roughly equivalent to a neighborhood, with 2,500-8,000 people per tract. These were the smallest/most granular way for us to collect and display data, which allowed us to visualize changes with the greatest detail. Within urban areas, the changing demographic patterns over time are often highly localized, down to even a few blocks. 

**Variables**

Initially, we searched for all of the relevant race-relevant variables in the 3 datasets (1990, 2000, 2010) in order to compare and determine what would best allow us to compare change over time. Each of the census years splits race a bit differently; in other words, race variables change significantly from one rollout of the census to the next. We ultimately chose to use, from each census year, the variable most closely corresponding to the number of people in each census tract who self-identified as white (regardless of those who marked Hispanic/Latinx ethnicity, which we would have liked to exclude from these numbers but were counted differently across the years). 

While our initial goal was to then expand and gather variables referring to other individual racial categories, we ultimately chose not to do this. While we recognize the value of collecting these data and displaying each racial category separately, we could not easily find continuity in these categories over time, which makes comparisons difficult. Further, we wanted each city's data to be easily displayed in a single map, and did not see a way to display numbers of several racial categories at a time without making the map overly complicated and uninterpretable. 

**Cities**

We decided to look at Boston and our hometowns to see if there were any trends in racial movement over the years.   

* Boston, MA - our current home!  
* District of Columbia - Erin's hometown!  
* Oakland/Bay Area, CA-  Kevin's hometown!  
* Knoxville, TN - Rachel's hometown!  

We found an article written by a collective of data scientists at Priceonomics titled, "How Diverse is Your City?" ("How Diverse is Your City", 2014).  The data scientists used ACS datasets to look into the diversity of the larger metropolitan areas of the nation. We chose some of the cities we found interesting from their findings.

* Seattle, WA - According to article, white people make up a large percentage of the population.    
* Portland, OR- Between 2000 and 2010, nearly 10,000 people of color moved out of the city. We wanted to see if we could visualize this.   
* Detroit, MI - The article mentioned that Detroit is a very diverse city, at 80.7% black when the article was written.   

**Visualizations/Dashboard**

We chose to display our visualizations in a dashboard because it allows for maximum interaction by the user, and allows for us to display a large number of visualizations in one place. 

Our dashboard allows for two types of views: one allowing the user to visualize individual maps, encouraging for examination of a single city's population racial distribution at a single point in time. The user can hover over any census tract for more information about the percent of the population that is white, and the number (magnitude) corresponding to that percent. 


The other view allows a user to compare two maps directly: within the same city, the user can compare the population's racial distribution at two different points in time. The scales and legend were adjusted to not change based on the data provided in order to ensure that the colors across years represented the same values for valid comparison when viewed visually. 

**Permutation Test**

We ran a permutation test only for the city of Boston. We decided to resample the years (i.e. 1990, 2000, 2010). **Null hypothesis:** racial distribution across areas of Boston has remained the same over time
See permutation tab of dashboard for the in-depth analysis and rationale

**References**

How Diverse is Your City? (2014). Retrieved May 12, 2019, from Priceonomics website: []http://priceonomics.com/how-diverse-is-your-city/



R Script
=====================================
Below is is the code we used to gather and prep the data, as well as perform the permutation tests and save the relevant datasets and objects

Column { vertical_layout: scroll}
-------------------------------------
```{r, eval=F, echo=T}
#### documentation ####
# Erin Mahoney, Rachel Kwon, Kevin Nazario
# S022, Spring 2019
# Final Project
# Topic: Racial Population Distribution in US Cities
# Note: See related RMD file to compile and run dashboard. 

#### set up ####
rm(list = ls())

library(magrittr)
library(texreg)
library(dplyr)
library(tidycensus)
library(ggplot2)
library(ggmap)
library(stringr)
library(tidyverse)
library(leaflet)
library(htmltools)
library(tigris)
library(acs)
library(gridExtra)
library(sf)


#### set variables ####
vars1990 <- c('P0070001') #all white population
vars2000 <- c('P003003') #all white population
vars2010 <- c('P003002') #all white population

#### BOSTON - Pulls and Cleaning ####
# set counties and pull data from census
countylist_MA <- c('Suffolk', 'Norfolk', 'Middlesex')

cendat_MA_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Massachusetts', county = countylist_MA)

cendat_MA_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Massachusetts', county = countylist_MA)

cendat_MA_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'Massachusetts', county = countylist_MA)

# Changing to %
cendat_MA_2010$pct_white <- round(cendat_MA_2010$P003002 / cendat_MA_2010$summary_value, 3) * 100 
cendat_MA_2000$pct_white <- round(cendat_MA_2000$P003003 / cendat_MA_2000$summary_value, 3) * 100 
cendat_MA_1990$pct_white <- round(cendat_MA_1990$P0070001 / cendat_MA_1990$summary_value, 3) * 100 

# Remove NAs from datasets
cendat_MA_2010 %<>% subset(!is.na(pct_white))
cendat_MA_2000 %<>% subset(!is.na(pct_white))
cendat_MA_1990 %<>% subset(!is.na(pct_white))

#### DC - Pulls and Cleaning ####
# pull data (no county necessary)
cendat_DC_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'District of Columbia')

cendat_DC_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'District of Columbia')

cendat_DC_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'District of Columbia')

# Change to %s
cendat_DC_2010$pct_white <- round(cendat_DC_2010$P003002 / cendat_DC_2010$summary_value, 3) * 100 
cendat_DC_2000$pct_white <- round(cendat_DC_2000$P003003 / cendat_DC_2000$summary_value, 3) * 100 
cendat_DC_1990$pct_white <- round(cendat_DC_1990$P0070001 / cendat_DC_1990$summary_value, 3) * 100 

# eliminate NAs
cendat_DC_2010 %<>% subset(!is.na(pct_white))
cendat_DC_2000 %<>% subset(!is.na(pct_white))
cendat_DC_1990 %<>% subset(!is.na(pct_white))

#### OAKLAND - Pulls and cleaning ####
# set counties and pull data from census
countylist_CA <- c('Alameda', 'Contra Costa', 'San Francisco', 'San Mateo', 'Marin', 'Santa Clara')

cendat_CA_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'California', county = countylist_CA)

cendat_CA_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'California', county = countylist_CA)

cendat_CA_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'California', county = countylist_CA)

# Changing to %
cendat_CA_2010$pct_white <- round(cendat_CA_2010$P003002 / cendat_CA_2010$summary_value, 3) * 100 
cendat_CA_2000$pct_white <- round(cendat_CA_2000$P003003 / cendat_CA_2000$summary_value, 3) * 100 
cendat_CA_1990$pct_white <- round(cendat_CA_1990$P0070001 / cendat_CA_1990$summary_value, 3) * 100 

# eliminate NAs
cendat_CA_2010 %<>% subset(!is.na(pct_white))
cendat_CA_2000 %<>% subset(!is.na(pct_white))
cendat_CA_1990 %<>% subset(!is.na(pct_white))

#### PORTLAND - Pulls and cleaning ####
# set counties and pull data from census
countylist_OR <- c('Columbia', 'Multnomah', 'Washington', 'Clackamas', 'Hood River')

cendat_OR_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Oregon', county = countylist_OR)

cendat_OR_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Oregon', county = countylist_OR)

cendat_OR_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'Oregon', county = countylist_OR)

# Changing to %
cendat_OR_2010$pct_white <- round(cendat_OR_2010$P003002 / cendat_OR_2010$summary_value, 3) * 100 
cendat_OR_2000$pct_white <- round(cendat_OR_2000$P003003 / cendat_OR_2000$summary_value, 3) * 100 
cendat_OR_1990$pct_white <- round(cendat_OR_1990$P0070001 / cendat_OR_1990$summary_value, 3) * 100 

# eliminate NAs
cendat_OR_2010 %<>% subset(!is.na(pct_white))
cendat_OR_2000 %<>% subset(!is.na(pct_white))
cendat_OR_1990 %<>% subset(!is.na(pct_white))

#### KNOXVILLE - Pulls and cleaning ####
# set counties and pull data from census
countylist_TN <- c('Knox', 'Anderson', 'Sevier', 'Blount')

cendat_TN_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Tennessee', county = countylist_TN)

cendat_TN_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Tennessee', county = countylist_TN)

cendat_TN_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'Tennessee', county = countylist_TN)

# Changing to %
cendat_TN_2010$pct_white <- round(cendat_TN_2010$P003002 / cendat_TN_2010$summary_value, 3) * 100 
cendat_TN_2000$pct_white <- round(cendat_TN_2000$P003003 / cendat_TN_2000$summary_value, 3) * 100 
cendat_TN_1990$pct_white <- round(cendat_TN_1990$P0070001 / cendat_TN_1990$summary_value, 3) * 100 

# eliminate NAs
cendat_TN_2010 %<>% subset(!is.na(pct_white))
cendat_TN_2000 %<>% subset(!is.na(pct_white))
cendat_TN_1990 %<>% subset(!is.na(pct_white))

#### SEATTLE - Pulls and Cleaning ####
# set counties and pull data from census
countylist_WA <- c('King', 'Snohomish', 'Kitsap', 'Island')

cendat_WA_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Washington', county = countylist_WA)

cendat_WA_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Washington', county = countylist_WA)

cendat_WA_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'Washington', county = countylist_WA)

# Changing to %
cendat_WA_2010$pct_white <- round(cendat_WA_2010$P003002 / cendat_WA_2010$summary_value, 3) * 100 
cendat_WA_2000$pct_white <- round(cendat_WA_2000$P003003 / cendat_WA_2000$summary_value, 3) * 100 
cendat_WA_1990$pct_white <- round(cendat_WA_1990$P0070001 / cendat_WA_1990$summary_value, 3) * 100 

# eliminate NAs
cendat_WA_2010 %<>% subset(!is.na(pct_white))
cendat_WA_2000 %<>% subset(!is.na(pct_white))
cendat_WA_1990 %<>% subset(!is.na(pct_white))

#### DETROIT - Pulls and Cleaning ####
# set counties and pull data from census
countylist_MI <- c('Oakland', 'Macomb', 'Wayne')

cendat_MI_2010 <- get_decennial(geography = "tract", variables = vars2010, year = 2010, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Michigan', county = countylist_MI)

cendat_MI_2000 <- get_decennial(geography = "tract", variables = vars2000, year = 2000, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P001001', state= 'Michigan', county = countylist_MI)

cendat_MI_1990 <- get_decennial(geography = "tract", variables = vars1990, year = 1990, geometry = TRUE, output = 'wide', shift_geo = FALSE, summary_var = 'P0010001', state= 'Michigan', county = countylist_MI)

# Changing to %
cendat_MI_2010$pct_white <- round(cendat_MI_2010$P003002 / cendat_MI_2010$summary_value, 3) * 100 
cendat_MI_2000$pct_white <- round(cendat_MI_2000$P003003 / cendat_MI_2000$summary_value, 3) * 100 
cendat_MI_1990$pct_white <- round(cendat_MI_1990$P0070001 / cendat_MI_1990$summary_value, 3) * 100 

# eliminate NAs
cendat_MI_2010 %<>% subset(!is.na(pct_white))
cendat_MI_2000 %<>% subset(!is.na(pct_white))
cendat_MI_1990 %<>% subset(!is.na(pct_white))

#### save datasets ####
save(cendat_MA_1990, cendat_MA_2000, cendat_MA_2010, cendat_DC_1990, cendat_DC_2000, cendat_DC_2010, cendat_CA_1990, cendat_CA_2000, cendat_CA_2010, cendat_OR_1990, cendat_OR_2000, cendat_OR_2010, cendat_WA_1990, cendat_WA_2000, cendat_WA_2010, cendat_TN_1990, cendat_TN_2000, cendat_TN_2010, cendat_MI_1990, cendat_MI_2000, cendat_MI_2010, file = 'C:/Users/Erin/Documents/GitHub/finalproject/SavedData.Rdata')


#### permutation files ####
# read csv files for year csv files. all these files are in your github path
year2000 <- read.csv("year2000.csv")
year1990 <- read.csv("year1990.csv")
year2010 <- read.csv("year2010.csv")

#### convert everything to dataframes for permutation tests-----------
year2000 <- as.data.frame(year2000)
year1990 <- as.data.frame(year1990)
year2010 <- as.data.frame(year2010)
cendat_MA_1990 <- as.data.frame(cendat_MA_1990)
cendat_MA_2000 <- as.data.frame(cendat_MA_2000)
cendat_MA_2010 <- as.data.frame(cendat_MA_2010)
#add years dataframe to census dataframe
testing_cendat_MA_2000 <- cbind(cendat_MA_2000,year2000)
testing_cendat_MA_1990 <- cbind(cendat_MA_1990,year1990)
testing_cendat_MA_2010 <- cbind(cendat_MA_2010,year2010)

#### subsetting dataframes for permutation testing-----------------------------------
sub_testing_cendat_MA_2000 <- as.data.frame(testing_cendat_MA_2000[,c('NAME','GEOID','year','pct_white')])
sub_testing_cendat_MA_2000 <- subset(sub_testing_cendat_MA_2000,select=c(NAME,GEOID,year,pct_white))
sub_testing_cendat_MA_1990 <- as.data.frame(testing_cendat_MA_1990[,c('NAME','GEOID','year','pct_white')])
sub_testing_cendat_MA_1990 <- subset(sub_testing_cendat_MA_1990,select=c(NAME,GEOID,year,pct_white))

sub_testing_cendat_MA_2010 <- as.data.frame(testing_cendat_MA_2010[,c('NAME','GEOID','year','pct_white')])
sub_testing_cendat_MA_2010 <- subset(sub_testing_cendat_MA_2010,select=c(NAME,GEOID,year,pct_white))


#### prepping for 3 permutations-----------------------------------------------------------
#merged 1990 and 2000 by year
testing_MA_1990_2000 <- rbind(sub_testing_cendat_MA_1990,sub_testing_cendat_MA_2000)
testing_MA_1990_2000$year %<>% factor(levels = c(1990, 2000), labels = c('1990', '2000'))
table(testing_MA_1990_2000$year)

#merged 2000 and 2010 by year
testing_MA_2000_2010 <- rbind(sub_testing_cendat_MA_2000,sub_testing_cendat_MA_2010)
testing_MA_2000_2010$year %<>% factor(levels = c(2000, 2010), labels = c('2000', '2010'))

#merged 1990 and 2010 by year
testing_MA_1990_2010 <- rbind(sub_testing_cendat_MA_1990,sub_testing_cendat_MA_2010)
testing_MA_1990_2010$year %<>% factor(levels = c(1990, 2010), labels = c('1990', '2010'))

#### for loop for 1990-2000 ------------------------------------------------------------------------
set.seed(1)
#testing
#resample year
testing_MA_1990_2000$year_permuted <- sample(testing_MA_1990_2000$year)
testing_MA_1990_2000[c('year', 'year_permuted')]

#ttest
t.test(pct_white ~ year_permuted, testing_MA_1990_2000)
R <- 10000
mean_diffs <- rep(NA, R)

#for loop for permuting
for(i in 1:R){
  testing_MA_1990_2000$year_permuted <- sample(testing_MA_1990_2000$year_permuted)
  means <- tapply(testing_MA_1990_2000$pct_white, testing_MA_1990_2000$year_permuted, mean,na.rm=TRUE)
  mean_diffs[i] <- means['2000'] - means['1990']
  if(i %% 1000 == 0) print(i)
}
means <- tapply(testing_MA_1990_2000$pct_white, testing_MA_1990_2000$year, mean,na.rm=TRUE)
observed_mean <- means['2000'] - means['1990']
mean_diffs <- data.frame(mean_diffs = mean_diffs)

#distibution of mean diffs 
plot1 <- ggplot(mean_diffs, aes(x = mean_diffs)) + geom_histogram() + geom_vline(xintercept = observed_mean, col = 'red', lty = 2) + ggtitle("Distribution of permuted mean differences: \nBoston 1990 & 2000 census")
#plot
plot1

#### permutation for 2000-2010-------------------------------------------------------
set.seed(1)
testing_MA_2000_2010$year_permuted <- sample(testing_MA_2000_2010$year)
testing_MA_2000_2010[c('year', 'year_permuted')]

# ttest
R <- 10000
mean_diffs <- rep(NA, R)

# for loop for permuting
for(i in 1:R){
  testing_MA_2000_2010$year_permuted <- sample(testing_MA_2000_2010$year_permuted)
  means <- tapply(testing_MA_2000_2010$pct_white, testing_MA_2000_2010$year_permuted, mean,na.rm=TRUE)
  mean_diffs[i] <- means['2010'] - means['2000']
  if(i %% 1000 == 0) print(i)
}
means <- tapply(testing_MA_2000_2010$pct_white, testing_MA_2000_2010$year, mean,na.rm=TRUE)
observed_mean <- means['2010'] - means['2000']
mean_diffs <- data.frame(mean_diffs = mean_diffs)
plot2 <- ggplot(mean_diffs, aes(x = mean_diffs)) + geom_histogram() + geom_vline(xintercept = observed_mean, col = 'red', lty = 2) + ggtitle("Distribution of permuted mean differences: \nBoston 2000 & 2010 census")
#plot
plot2

#### permutation for 1990-2010-------------------------------------------------------
set.seed(1)
testing_MA_1990_2010$year_permuted <- sample(testing_MA_1990_2010$year)
testing_MA_1990_2010[c('year', 'year_permuted')]

# ttest
t.test(pct_white ~ year_permuted, testing_MA_1990_2010)
R <- 10000
mean_diffs <- rep(NA, R)

# for loop for permuting
for(i in 1:R){
  testing_MA_1990_2010$year_permuted <- sample(testing_MA_1990_2010$year_permuted)
  means <- tapply(testing_MA_1990_2010$pct_white, testing_MA_1990_2010$year_permuted, mean,na.rm=TRUE)
  mean_diffs[i] <- means['2010'] - means['1990']
  if(i %% 1000 == 0) print(i)
}
means <- tapply(testing_MA_1990_2010$pct_white, testing_MA_1990_2010$year, mean,na.rm=TRUE)
observed_mean <- means['2010'] - means['1990']
mean_diffs <- data.frame(mean_diffs = mean_diffs)
plot3 <- ggplot(mean_diffs, aes(x = mean_diffs)) + geom_histogram() + geom_vline(xintercept = observed_mean, col = 'red', lty = 2) + ggtitle("Distribution of permuted mean differences: \nBoston 1990 & 2010 census")
#plot
plot3

#### save permutation plots ####
save(plot1, plot2, plot3, file = 'C:/Users/Erin/Documents/GitHub/finalproject/SavedPlots.Rdata')

#### consolidate saved file ####
rm(list = ls())

load("SavedData.Rdata")
load("SavedPlots.Rdata")

save(cendat_MA_1990, cendat_MA_2000, cendat_MA_2010, cendat_DC_1990, cendat_DC_2000, cendat_DC_2010, cendat_CA_1990, cendat_CA_2000, cendat_CA_2010, cendat_OR_1990, cendat_OR_2000, cendat_OR_2010, cendat_WA_1990, cendat_WA_2000, cendat_WA_2010, cendat_TN_1990, cendat_TN_2000, cendat_TN_2010, cendat_MI_1990, cendat_MI_2000, cendat_MI_2010, plot1, plot2, plot3, file = 'C:/Users/Erin/Documents/GitHub/finalproject/Saved.Rdata')
```
